Individual-based numerical experiment to describe the distribution of floating kelp within the Southern Benguela Upwelling System.

### 2022-08-31 thesis_chapter_4.Rmd
uploaded updated analysis files

---
title: "Individual-based numerical experiment to describe the distribution of floating kelp within the Southern Benguela Upwelling System."
author: 
- R Coppin^[Department of Biodiversity and Conservation Biology, University of the Western Cape, Private Bag X17, Bellville 7535, South Africa], C Rautenbach^[National Institute of Water and Atmospheric Research, Hamilton: Gate 10 Silverdale Road Hillcrest, 3216], AJ Smit^[South African Environmental Observation Network, Elwandle Coastal Node, Port Elizabeth, South Africa]
output:
  word_document: 
    reference_docx: markdown/wiley_template.docx
  html_document: default
  pdf_document: default
bibliography: references.bib
csl: markdown/limnologyandoceanography.csl
---

@@

The inter- to sub-tidal habitats of temperate shorelines are usually dominated by species of macroalgae, specifically brown seaweeds of the order Laminariales; commonly referred to as kelp [@Steneck2002;@Schiel2015-vr]. Kelps are highly productive organisms [@Mann1973;@Miller2011-pm; @Dayton1998-xw;@Krumhansl2012;@Pedersen2012-vr] which provide a 3-dimensional habitat for various fauna and flora [@anderson1997;@Leliaert2000a;@Steneck2002; @Norderhaug2012-ks; @Krumhansl2016]. Kelp are also capable of altering the local wave and light environment by wave dampening [@Mork1996-md; @Mia2013] and shading of the understorey habitat [@duggins1990; @watanabe1992; @clark2004], as well as providing important organic subsidy through erosion of fronds and fragmentation of plant structures [@Krumhansl2012; @Filbee-Dexter2018-wp; @DeBettignies2013; @Simonson2015; @ramirez2021]. Due to kelps’ ecological role in the nearshore, they are regarded as ecological engineers as well as important influencer of biodiversity along temperate shorelines [@FowlerWalker2005;@graham2004;@Harley2012;@Smale2017; @Wing2007-tv;@Wernberg2008]. Kelps are also important from a socio-economic perspective, as they provide habitats for economically important species such as abalone and rock lobster [@vasquez2014; @blamey2018], and are harvested for uses ranging from feed for the aquaculture sector [@troell2006] to pharmaceuticals [@smit2004].

Kelps’ ecological and socio-economic contributions are not limited to their local environment. During times of high wave energy kelps can become dislodged from the substrate  [@seymour1989; @Thomsen2004] and are transported via ocean currents, eventually deposited on beaches or offshore areas [@hobday2000a]. The kelp deposited on beaches provides important organic subsidies for coastal invertebrate communities, as well as a resource which can be harvested for economic purposes [@kirkman1997;@troell2006; @holden2018]. Along with the organic input are 'hitchhikers', which are organisms that remain attached during offshore transport once dislodged [@highsmith1985; @holmquist1994; @macaya2016; @thiel2005]. The transport of other organisms attached to kelp is regarded as an important processes in invasive species intrusion [@lewis2005; @McCormick2008; @nikula2010; @batista2018] and marine connectivity [@norton1992, @ingolfsson1995; @hobday2000a; @fraser2011]. The formation of kelp-rafts can also occur, providing habitat complexity in the pelagic environment for various fish species [@kingsford1992; @kingsford1995], Kelps which are transported offshore can become entangled to form a kelp-raft, but there is no clear mechanism for kelp-raft formation. The structural integrity of individual kelp or rafts will degrade over time through a combination of biotic and abiotic factors which ultimately results in sinking. Once settled on the seafloor, the kelp will degrade further over time resulting in organic input into offshore areas; which may be transported back towards the coast through upwelling processes [@Krumhansl2012; @Filbee-Dexter2018-wp; @ramirez2021].

Understanding the distribution patterns of macroalgae in the ocean and the underlying oceanographic processes influencing trajectory has received much attention in the literature. Approaches such as radio-tagged kelp rafts [@harrold1989], geological forms in holdfasts [@waters2018], genetic approaches [@craw2018; @waters2018; @nikula2013], the presence of region-specific epibionts on beach-cast kelp [@fraser2011], and Lagrangian approaches [@putman2018; @brooks2019; @putman2020] have helped to shed light on the processes involved in macroalgae transport.  Ocean currents are the main influencer of trajectory as well as drag and windage, which is the direct effect of wind on a floating object [@onink2019; @shen2019; @dasaro2018; @gates2018; @beron2016; @trinanes2016; @furnans2008]. More recent work by @coppin2021 has highlighted the importance of both hydrodynamic and wind drag in entrainment patterns within oceanographic mesoscale processes, more specifically eddies.

Although the past research concerning the transport of kelp has greatly improved our understanding of such processes, limitations and gaps still exist. For example, kelps are highly morphologically diverse which may influence buoyancy and overall drag characteristics. Morphological characteristics, sea surface temperatures, and light also influence the time which kelps remain afloat at the surface; known as raft-times. Macroalgae rafts also vary greatly in size, therefore studies investigating a specific raft size limits our understanding of that particular scenario as the findings cannot be inferred for other sizes of kelp rafts. Finally, it has been well established that morphological characteristics of macroalgae are influenced by local abiotic factors, which can vary greatly along coastlines. Although factors such as light, turbidity, and nutrients have been shown to influence macroalgae morphology; temperature and waves are regarded as the main influencers [@Hurd2000;@Denny2002a;@Thomsen2004;@FowlerWalker2005;@Wernberg2005c;@mabin2013;@bekkby2014; @coppin2021]. How the variable response in morphological characteristics influences sink rates, buoyancy, and drag is also yet to be established. There is also no current work where rafts may form on the surface and the role of eddies in this regard. It is most likely that eddies might play a role in this regard as they have shown to cause collection of other flotsam at the surface via entrainment @onink2019; @dasaro2018; @brach2018. For example, a study by @beron2019 showed that the inclusion of inertia in numerical experiments resulted in entrainment patterns of flotsam in eddies which more closely reflected observational data.

A better understanding of the distribution of floating kelp and ocean processes involved will aid in understanding the broader ecological role that kelp plays in the marine environment but will also be useful for coastal management and harvesting activities. The current work aims to shed light on the distribution patterns of floating kelp by a series of Lagrangian individual-particle tracking numerical experiments. This approach will allow for the morphological characteristics that vary with the local temperature and wave environment to be accounted for. In addition, since individual kelp has a higher probability of dislodgement compared to kelps in which holdfasts are clustered together, an individual-based approach will reflect real-world scenarios [@Thomsen2004]. The novel approach of including hydrodynamic and wind drag based on site-specific morphological characteristics by @coppin2021 will be used and the current work will build on this approach by the inclusion of morphological specific buoyancy and sinking rates.

@@

@@

Morphological characteristics of kelps are influenced by local wave and temperature conditions, which in turn will influence buoyancy, drag, and sinking parameters. Kelp morphological data was garnered from a previous study by @coppin2020. The assumption made is that the largest kelps (largest surface area) are the most likely to be dislodged from the substrate, therefore only maximum surface areas were garnered from the data.  In instances where the morphological data needed were not available, reasonable estimates were used based on previous studies and experience. To calculate the surface area and volume for kelps from each site, geometric shapes were used which most closely matched the shape of the relevant kelp morphological characteristics as per @coppin2021. The shapes assigned to each of the morphological features were used in the calculation of buoyancy, drag, and sinking rate.

@@

To effectively model the trajectory of a floating object, buoyancy needs to be calculated. The buoyancy will determine the amount of exposure to, and ultimately magnitude, of hydrodynamic and wind drag which in turn influences the trajectory of the object to the variable current and wind fields. The portion of an object above and below the surface of the water is determined by the net effect of forces acting on the object, namely buoyancy force (BF) and the weight of the object. The BF is the force pushing the object out of the water and the weight is the force pulling the object down. In order to calculate the buoyancy force (BF) the density and volume of the object must be known, such that: $$BF = p_fV_sg$$

@@

The solution of equation (3) results in a unitless fraction that can be expressed as a percentage of the relevant portion submerged. This in turn can be used to calculate the percentage of the object above the surface of the water (exposed to wind). The percentage of the object below and above the water surface was used in the calculation of hydrodynamic and wind drag when the virtual kelp is at the surface. Since morphological characteristics are influenced by the local wave and temperature environment, the buoyancy of virtual kelp particles may vary from site to site.

@@

Kelp floating on the surface of the ocean will begin to sink once the structural integrity of air-filled structures of the plants fails and becomes flooded. It is also possible for sinking to begin due to an increase in epibiont biomass which overwhelms the buoyancy of kelp, however, this is beyond the scope of this study. The target species of this study, *E. maxima*, contains a hollow stipe and a single air-filled pneumatocyst which serve to keep the plant upright in the water column, allowing the light-saturated surface waters to be reached. These structures also cause the plant to float if dislodged and therefore was taken into account when calculating the overall volume of the plant; which in turn was used to calculate other important parameters (buoyancy, drag, and sink rate).

## Vertical Velocity

@@

Where $p_s$ is the density of the particle, $p__l$ the density of the medium, $g$ is the gravitational acceleration constant, and $V$ is the total volume of the particle. In order to accurately calculate the density of the kelp based on volume and mass, the volume of the hollow section of the stipe was subtracted from the overall volume. The target species used for this experiment (*E. maxima*) has an air-filled stipe and therefore the hollowness needs to be accounted for when calculating the overall volume of the stipe. The volume of the kelp is:
$$Volume = V_t - V_h$$
where $V_t$ is the total volume of the stipe if now hollowness is taken into account and $V_h$ is the hollowness volume of the stipe.
The other component of the vertical velocity is the drag force in the opposite direction to the downward force of the weight. The vertical drag component is calculated by the same equation for drag in the horizontal, such that:

@@

Simulations used daily outputs of velocity vectors (meridonal and zonal) and salinity (psu) for varying depths for the study region. To account for the role of morphological characteristics in influencing drift trajectory in terms of hydrodynamic and wind drag [@coppin2021] simulations were run separately for each site per season. Morphological characteristics of macroalgae has been shown to be based on the local wave and temperature environment, hence the need for site specific parameters with regards to morphological characteristics. Running simulations per season allowed incorporation of seasonal raft-times identified in pervious work [@graiff2016 and @graiff2013]. The outputs from each simulation (site) was pooled based on season.

Simulations were run using the OceanParcels software (available from https://oceanparcels.org/). All simulations were run using a Runge-Kutte fourth order advection scheme, while custom kernels simulated the effects of hydrodynamic and wind drag on overall trajectory, seasonal raft-times, sinking and settling on seafloor of virtual kelp particles; all important processes in the determination of surface and bottom sinks. The simulations were run for 2019, which was divided into various seasons (see table 2). For incorporating drag, the site specific buoyancy calculations were used to determine the magnitude of hydrodynamic (portion below surface) and wind (portion above the surface) while virtual kelp particles were on the surface and then converted to full hydrodynamic drag exposure (100%) once the virtual kelp started sinking; as wind would no longer act as a direct influencer of trajectory once below the ocean surface.

@@

Raft-time or sometimes referred to as raft-longevity is the time a macroalgae raft remains afloat at the surface of the ocean after detachment before sinking. Raft-time, also referred to as raft longevity, is determined by epibiont load, macroalgal growth rate and the loss of kelp biomass over time through fragmentation. Epibiont load consists of epiphytic and bryozoan species which tend to grow more rapidly in higher light and temperature environments, such as the ocean surface [@thiel2005; @thiel2005; @graiff2016]. The increase in epibiont load reduces the buoyancy of floating macroalgal rafts over time eventually causing them to sink to the seafloor, while also potentially increasing drag [@hobday2000; @tala2013; @craw2018]. Fragmentation of kelp biomass over time may also reduce raft times further, while growth may counteract the effects of epibiont load and fragmentation provided the environmental conditions are favorable [@graiff2016; @macaya2016]. For instance a study by @rothausler2011 investigated the abiotic and biotic factors influencing raft time and dispersal potential of the giant kelp *M. pyrifera* by a combination of tethering experiments and field surveys. The results showed the physiological performance of kelp declined with increasing epiphyte biomass and that higher temperatures increased growth rate of epibionts and fragmentation of kelp. The authors concluded from the results of both the experiments and field surveys that *M. pyrifera* dispersal is dependent on low temperature and moderate irradiance conditions, with high temperatures and higher irradiance reducing overall raft-time and hence dispersal potential [@rothausler2011]. Raft-time was estimated from the results of past studies by @graiff2016 and @graiff2013 for each season. For summer the raft time was 22 days, autumn was 41 days, winter was 38 days, and spring was 33 days.

@@

```{r morphology surface area table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE}
site_surface_area <- set_caption(site_surface_area, "Table 1:Summary of the attributes and estimates used to calculate the overall site-specific surface area of kelp individuals used in the various simulations.") 
#site_surface_area <- set_table_properties(site_surface_area, layout = "autofit")
site_surface_area

Simulations used daily outputs of velocity vectors (meridonal and zonal) and salinity (psu) for varying depths for the study region. To account for the role of morphological characteristics in influencing drift trajectory in terms of hydrodynamic and wind drag [@coppin2021] simulations were run separately for each site per season. Morphological characteristics of macroalgae has been shown to be based on the local wave and temperature environment, hence the need for site specific parameters with regards to morphological characteristics. Running simulations per season allowed incorporation of seasonal raft-times identified in pervious work [@graiff2016 and @graiff2013]. The outputs from each simulation (site) was pooled based on season.

Simulations were run using the OceanParcels software (available from https://oceanparcels.org/). All simulations were run using a Runge-Kutte fourth order advection scheme, while custom kernels simulated the effects of hydrodynamic and wind drag on overall trajectory, seasonal raft-times, sinking and settling on seafloor of virtual kelp particles; all important processes in the determination of surface and bottom sinks. The simulations were run for two years, which were divided into the relevant seasons seasons (see table 2).

```{r simulation table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE, fig.cap="TBD."}

chapter4_table2 <- read_csv2(file = "tables/chapter_4/simulation_table.csv")
chapter4_table2 <- flextable(data = chapter4_table2)
chapter4_table2 <- theme_zebra(chapter4_table2, odd_header = "transparent", even_header = "transparent")
chapter4_table2 <- align(x= chapter4_table2, align = "center", part = "all")
chapter4_table2 <- theme_vanilla(chapter4_table2)
chapter4_table2 <- fontsize(chapter4_table2, size = 8, part = "body")
chapter4_table2 <- fontsize(chapter4_table2, size = 8, part = "header")
chapter4_table2 <- colformat_num(chapter4_table2,digits = 1)
chapter4_table2 <- set_table_properties(chapter4_table2, layout = "autofit")
chapter4_table2 <- set_caption(chapter4_table2, "Table 2: Summary of the simulations run for the study. Simulations were run for each site over the course of 2 years and then pooled based on season for each year. The details pertaining are provided in the below. The start of each year was the beginning of Austral summer.") 
chapter4_table2
```

@@

To compare virtual kelp trajectories between sites the mean trajectories were calculated, and density distribution maps were produced. Density distribution maps are an effective method for assessing pathways and ocean connectivity [@vanSebille2018]. The distance particles traveled was also measured and used to produce boxplots. Distance traveled from the release site can help reveal topographic steering and momentum energy transfer from the wind field to kelp particles [@vanSebille2018]. If the wind direction is in the same direction as the current the virtual particles will cover larger distances due to an increase in the velocity because of momentum energy transfer [@putman2020; @putman2018; @vanSebille2018]. To aid in seasonal comparison of trajectories, proportions of each kelp phase at the end of each simulation run was calculated and compared. The different phases were surface, subsurface, beached, sunk and out of bounds.

>>>>

---
title: 'Individual-based numerical experiment to describe the distribution of floating kelp within the Southern Benguela Upwelling System.'
author: 
- R Coppin^[Department of Biodiversity and Conservation Biology, University of the Western Cape, Private Bag X17, Bellville 7535, South Africa]
- AJ Smit^[Department of Biodiversity and Conservation Biology, University of the Western Cape, Private Bag X17, Bellville 7535, South Africa, South African Environmental Observation Network, Elwandle Coastal Node, Port Elizabeth, South Africa]
- C Rautenbach^[National Institute of Water and Atmospheric Research, Hamilton: Gate 10 Silverdale Road Hillcrest, 3216]
documentclass: article
linestretch: 1.5
fontfamily: accanthis #accanthis #txfonts
pagestyle: plain
geometry: "left=2.5cm,right=2.5cm,top=2cm,bottom=2cm"
indent: false
output:
  word_document:
    reference_docx: markdown/wiley_template.docx
  html_document:
  pdf_document:
    latex_engine: xelatex
    toc: no
bibliography: references.bib
csl: markdown/limnologyandoceanography.csl
---

@@

The rocky inter- to sub-tidal habitats of temperate shorelines are usually dominated by species of macroalgae, specifically brown seaweeds of the order Laminariales; commonly referred to as kelp [@Steneck2002;@Schiel2015-vr]. Kelps are highly productive organisms [@Mann1973;@Miller2011-pm; @Dayton1998-xw;@Krumhansl2012;@Pedersen2012a] which provide a 3-dimensional habitat for various fauna and flora [@anderson1997;@Leliaert2000a;@Steneck2002; @Norderhaug2012-ks; @Krumhansl2016]. Kelp are also capable of altering the local wave and light environment by wave dampening [@Mork1996-md; @Mia2013] and shading of the understorey habitat [@duggins1990; @watanabe1992; @clark2004], as well as providing important organic subsidy through erosion of fronds and fragmentation of plant structures [@Krumhansl2012; @filbee2018; @DeBettignies2013; @Simonson2015; @ramirez2021]. Due to kelps’ ecological role in the nearshore, they are regarded as ecological engineers as well as important influencer of biodiversity along temperate shorelines [@FowlerWalker2005;@graham2004;@Harley2012;@Smale2017; @Wing2007-tv;@Wernberg2008]. Kelps are also important from a socio-economic perspective, as they provide habitats for economically important species such as abalone and rock lobster [@vasquez2014; @blamey2018], and are harvested for uses ranging from feed for the aquaculture sector [@troell2006] to pharmaceuticals [@smit2004].

Kelps’ ecological and socio-economic contributions are not limited to their local environment. During times of high wave energy kelps can become dislodged from the substrate [@seymour1989; @Thomsen2004] and are transported via ocean currents, eventually deposited on beaches or offshore areas [@hobday2000a]. The kelp deposited on beaches provides important organic subsidies for coastal invertebrate communities [@griffiths1983; @michaud2019; @gilson2021], as well as a resource which can be harvested for economic purposes [@kirkman1997;@troell2006; @holden2018]. Along with the organic input are 'hitchhikers', which are organisms that remain attached during offshore transport once dislodged [@highsmith1985; @holmquist1994; @macaya2016; @thiel2005]. The transport of other organisms attached to kelp is regarded as an important processes in invasive species intrusion [@lewis2005; @McCormick2008; @nikula2010; @batista2018] and marine connectivity [@norton1992, @ingolfsson1995; @hobday2000a; @fraser2011]. Kelps which are transported offshore can become entangled to form a kelp-raft, providing habitat complexity in the pelagic environment for various fish species [@kingsford1992; @kingsford1995]. However, there is no clear mechanism for kelp-raft formation. The structural integrity of individual kelp or rafts will degrade over time through a combination of biotic and abiotic factors which ultimately results in sinking [@graiff2016; @macaya2016; @smith2002; @thiel2005]. Once settled on the seafloor, the kelp will degrade further over time resulting in organic input into offshore areas; which may be transported back towards the coast through upwelling processes [@Krumhansl2012; @filbee2018; @ramirez2021].

Understanding the distribution patterns of macroalgae in the ocean and the underlying oceanographic processes influencing trajectory has received much attention in the literature. Approaches such as radio-tagged kelp rafts [@harrold1989], geological forms in holdfasts [@waters2018], genetic approaches [@craw2018; @waters2018; @nikula2013], the presence of region-specific epibionts on beach-cast kelp [@fraser2011], and Lagrangian approaches [@putman2018; @brooks2019; @putman2020] have helped to shed light on the processes involved in macroalgae transport.  Ocean currents are the main influencer of trajectory as well as drag and windage, which is the direct effect of wind on a floating object [@onink2019; @shen2019; @dasaro2018; @gates2018; @beron2016; @trinanes2016; @furnans2008]. More recent work by @coppin2022 has highlighted the importance of both hydrodynamic and wind drag in entrainment patterns within oceanographic mesoscale processes, more specifically eddies.

Although the past research concerning the transport of kelp has greatly improved our understanding of such processes, limitations and gaps still exist. For example, kelps are highly morphologically diverse which may influence buoyancy and overall drag characteristics. Morphological characteristics, sea surface temperatures, and light also influence the time which kelps remain afloat at the surface; known as raft-time. Macroalgae rafts also vary greatly in size, therefore studies investigating a specific raft size limits our understanding of that particular scenario as the findings cannot be inferred for other sizes of kelp rafts. Finally, it has been well established that morphological characteristics of macroalgae are influenced by local abiotic factors, which can vary greatly along coastlines. Although factors such as light, turbidity, and nutrients have been shown to influence macroalgae morphology; temperature and waves are regarded as the main influencers [@Hurd2000;@Denny2002a;@Thomsen2004;@FowlerWalker2005;@Wernberg2005c;@mabin2013;@bekkby2014; @coppin2022]. How the variable response in morphological characteristics influences sink rates, buoyancy, and drag is also yet to be established. There is also no current work where rafts may form on the surface. It is most likely that eddies might play a role in this regard as they have shown to cause collection of other flotsam at the surface via entrainment [@onink2019; @dasaro2018; @brach2018]. For example, a study by @beron2019 showed that the inclusion of inertia in numerical experiments resulted in entrainment patterns of flotsam in eddies which more closely reflected in observational data.

A better understanding of the distribution of floating kelp and ocean processes involved will aid in understanding the broader ecological role that kelp plays in the marine environment but will also be useful for coastal management and harvesting activities. The current work aims to shed light on the distribution patterns of floating kelp by a series of Lagrangian individual-particle tracking numerical experiments. This approach will allow for the morphological characteristics that vary with the local temperature and wave environment to be accounted for. In addition, since individual kelp has a higher probability of dislodgement compared to kelps in which holdfasts are clustered together, an individual-based approach will reflect real-world scenarios [@Thomsen2004]. The approach of including hydrodynamic and wind drag based on site-specific morphological characteristics by @coppin2022 will be used and the current work will build on this approach by the inclusion of morphological specific buoyancy and sinking rates.

@@

## Study site

The region of interest for this study was the Cape Peninsula in the Western Cape (Figure 1A). Kommetjie, located on the Atlantic side of the Cape Peninsula (Figure 2B), was chosen as a release site. The hydrodynamics in the study region are characterized by a complex interaction of wind, sea waves and ocean swell, and are also influenced by larger-scale regional oceanic processes. The larger processes in the region can be attributed to the Agulhas Current (AC) and the Benguela Current (BC) [@lutjeharms2000;@lutjeharms2007;@garzoli1996;@hutchings2009;@rubio2009]. The AC is a component of the South-west Indian Ocean sub-gyre, flowing predominantly south-westward following the continental shelf edge and eventually retroflecting eastward back into the South Indian Ocean [@lutjeharms2000;@lutjeharms2007].

The region where the retroflection occurs is known as the Agulhas Retroflection and consists of high levels of mesoscale variability [@garzoli1996]. The variability results in shedding of rings and filaments known as the Agulhas Rings and have an average diameter of ~324km [@lutjeharms1988;@lutjeharms2006; @lutjeharms2007]. Agulhas Rings are the main contributors to the Agulhas leakage which is the transport of warm water from the Indian Ocean to the South Atlantic Ocean [@garzoli1996; @lutjeharms2006; @beal2015]. The Agulhas leakage is also a source of meso-scale variability within the Benguela region as anticyclonic rings are shed and drift northwestward along the southern Benguela slope region [@rubio2009]. The Benguela region ---known as the Benguela Upwelling System (BUS)--- is characterized by strong coastal upwelling which is primarily driven by equatorward winds [@shannon1996;@hardman2003;@veitch2018; @veitch2017]. The BUS starts at 27$^\circ$S and extends southward to 35$^\circ$S [@shannon1996]. The region is also characterized by large amounts of mesoscale variability in the form of eddys and filaments [@blanke2002; @blanke2005; @rubio2009]. Inshore, the BUS is dominated by a northwestward flow, known as the Benguela Jet, due to topographical steering, wind stress and interactions with passing Agulhas Rings and eddys [@stramma1989; @veitch2010; @veitch2017; @veitch2018; @hutchings2009]. The stability of the jet varies seasonally with more stable flows in winter; in January, the jet begins to split between Cape Point and Cape Columbine [@veitch2018; @blanke2005; @blanke2002] which results in a weak narrow inshore branch that flows along the shelf edge and a strong flowing north-westwards offshore branch. The area is also characterized by high upwelling activity driven by strong westerly winds [@hutchings2009]. These winds are most intense during Austral summer (December-March) which favours upwelling conditions in the region, and also act to reinforce the flow of the Benguela Jet which favours upwelling waters contributing to the across-shore density gradient [@ragoasha2019]. In Austral winter the wind direction is reversed which decreases upwelling favorable conditions, but the core of the Benguela Jet is maintained by the intrusion of warm Agulhas waters into the Benguela region over the shelf edge [@veitch2018; @hutchings2009; @ragoasha2019].

```{r map chp4, echo=FALSE, message=FALSE, warning=FALSE, strip.white=TRUE, fig.cap="Map of the study domain and the main features pertaining to transport in the region. In panel A, the map of the western portion of the South African coastline and the main currents is presented. The resulting eddies and filaments forming from the Agulhas retroflection are indicated, and the position of the study area is drawn in as a black rectangle. Details of the study area are given in panel B. Here the splitting of the Benguela Jet, eddy features, and the general flow within the study domain are presented. The release site of simulated kelp ‘particles’ is represented by a black triangle."}

library(png)
library(grid)
library(gridExtra)

#map_chp3A <- rasterGrob(as.raster(readPNG("map/chp3_mapA.png")))
#map_chp3B <- rasterGrob(as.raster(readPNG("map/chp3_mapB.png")))
#ggarrange(map_chp3A, map_chp3B, ncol = 2, nrow = 1, labels = "AUTO", align = c("h"))
map_chp3_combined <- rasterGrob(as.raster(readPNG("map/chp3_map_final_v2.png")))
ggarrange(map_chp3_combined)

Simulations used daily outputs of velocity vectors (meridonal and zonal) and salinity (psu) for varying depths for the study region. To account for the role of morphological characteristics in influencing drift trajectory in terms of hydrodynamic and wind drag [@coppin2022] simulations were run separately for each site per season. Morphological characteristics of macroalgae has been shown to be based on the local wave and temperature environment, hence the need for site specific parameters with regards to morphological characteristics. Running simulations per season allowed incorporation of seasonal raft-times identified in pervious work [@graiff2016 and @graiff2013]. The outputs from each simulation (site) was pooled based on season.

Simulations were run using the OceanParcels software (available from https://oceanparcels.org/). All simulations were run using a Runge-Kutte fourth order advection scheme, while custom kernels simulated the effects of hydrodynamic and wind drag on overall trajectory, seasonal raft-times, sinking and settling on seafloor of virtual kelp particles; all important processes in the determination of surface and bottom sinks. The simulations were run for two years, which were divided into the relevant seasons (see table 2). For incorporating drag, the site specific buoyancy calculations were used to determine the magnitude of hydrodynamic (portion below surface) and wind (portion above the surface) while virtual kelp particles were on the surface and then converted to full hydrodynamic drag exposure (100%) once the virtual kelp started sinking; as wind would no longer act as a direct influencer of trajectory once below the ocean surface.

Although OceanParcels is a robust and flexible tool, it does have some limitations when combining mixed field of wind and ocean vectors with sinking behaviour. Essentially, problems arise when including wind fields with ocean vector fields when simulations include sinking. This creates limitations when running simulations and requires the approach to be adapted. It was decided to incorporate an approach used in previous work by Putman et al. 2018, Putman et al. 2020, need more here, such that windage is estimated as 50% of the current speed. Therefore, the total potential wind was estimated as 50% of the current speed to attain the velocity vectors for the estimated wind field, which was then used to calculate the total effect of windage based on buoyancy characteristics (magnitude of surface area exposed to wind and hydrodynamic drag respectively.

Kelp floating on the surface of the ocean will begin to sink once the structural integrity of air-filled structures of the plants fails and becomes flooded. It is also possible for sinking to begin due to an increase in epibiont biomass which overwhelms the buoyancy of kelp, however, this is beyond the scope of this study. The target species of this study, *Ecklonia maxima* (Osbeck) Papenfuss, contains a hollow stipe and a single air-filled pneumatocyst which acts to keep the plant upright in the water column, allowing the light-saturated surface waters to be reached. These structures also cause the plant to float if dislodged and therefore were taken into account when calculating the overall volume of the plant; which in turn was used to calculate other important parameters (buoyancy, drag, and sink rate).

@@

Morphological characteristics of kelps are influenced by local wave and temperature conditions, which in turn will influence buoyancy, drag, and sinking parameters. Kelp morphological data was garnered from a previous study by @coppin2020. The assumption made is that the largest kelps (largest surface area) are the most likely to be dislodged from the substrate, therefore only maximum surface areas were garnered from the data. In instances where the morphological data needed were not available, reasonable estimates were used based on previous studies and experience.

The exception was the air-filled stipe, which helps to keep the plant upright within the water column, where ad-hoc fieldwork was performed to attain better estimates of the amount of air within the stipe. The hollowness of E. maxima has been investigated as a part of a previously by Rothman et al. (2017) who showed that the relative hollowness did not significantly differ among sites, with 86-90% of the total stipe length containing air. Therefore, based on these results, it is assumed that *E. maxima* individuals will exhibit the same degree of hollowness amongst populations along South African coastline. An estimate of stipe thickness of the hollow sections was used in conjunction with the 90% of the stipe length shown by @Rothman2017-pn to calculate a reasonable estimate for the volume of air in the stipe. This is an important aspect as it directly effects the buoyancy of an individual floating kelp which ultimately determines the magnitude of exposure to hydrodynamic and wind drag.

To calculate the surface area and volume for kelps from each site, geometric shapes were used which most closely matched the shape of the relevant kelp morphological characteristics as per @coppin2022. The shapes assigned to each of the morphological features were used in the calculation of buoyancy, drag, and sinking rate.

@@

To effectively model the trajectory of a floating object, buoyancy needs to be calculated. The buoyancy will determine the amount of exposure to, and ultimately magnitude, of hydrodynamic and wind drag which in turn influences the trajectory of the object to the variable current and wind fields [@dietrich1982; @lobelle2021; @kooi2017]. The portion of an object above and below the surface of the water is determined by the net effect of forces acting on the object, namely buoyancy force (BF) and the weight of the object[@dietrich1982; @lobelle2021; @kooi2017]. The BF is the force pushing the object out of the water and the weight is the force pulling the object down. In order to calculate the buoyancy force (BF) the density and volume of the object must be known, such that: $$BF = p_fV_sg$$

@@

The solution of equation (3) results in a unitless fraction that can be expressed as a percentage of the relevant portion submerged. This in turn can be used to calculate the percentage of the object above the surface of the water (exposed to wind). The percentage of the object below and above the water surface was used in the calculation of hydrodynamic and wind drag when the virtual kelp is at the surface, while only hydrodynamic drag is considered for horizontal transport when sinking. Since morphological characteristics are influenced by the local wave and temperature environment, the buoyancy of virtual kelp particles may vary from site to site.

@@

@@

Where $p_s$ is the density of the particle, $p_l$ the density of the medium, $g$ is the gravitational acceleration constant, and $V$ is the total volume of the particle. In order to accurately calculate the density of the kelp based on volume and mass, the volume of the hollow section of the stipe was subtracted from the overall volume. The target species used for this experiment (*E. maxima*) has an air-filled stipe and therefore the hollowness needs to be accounted for when calculating the overall volume of the stipe. The volume of the kelp is:
$$Volume = V_t - V_h$$
where $V_t$ is the total volume of the stipe if now hollowness is taken into account and $V_h$ is the hollowness volume of the stipe.
The other component of the vertical velocity is the drag force in the opposite direction to the downward force of the weight. The vertical drag component is calculated by the same equation for drag in the horizontal, such that:

@@

@@

Raft-time or sometimes referred to as raft-longevity is the time a macroalgae raft remains afloat at the surface of the ocean after detachment before sinking. Raft-time, also referred to as raft longevity, is determined by epibiont load, macroalgal growth rate and the loss of kelp biomass over time through fragmentation. Epibiont load consists of epiphytic and bryozoan species which tend to grow more rapidly in higher light and temperature environments, such as the ocean surface [@smith2002; @thiel2005; @graiff2016]. The increase in epibiont load reduces the buoyancy of floating macroalgal rafts over time eventually causing them to sink to the seafloor, while also potentially increasing drag [@hobday2000; @tala2013; @craw2018]. Fragmentation of kelp biomass over time may also reduce raft times further, while growth may counteract the effects of epibiont load and fragmentation provided the environmental conditions are favorable [@graiff2016; @macaya2016]. For instance a study by @rothausler2011 investigated the abiotic and biotic factors influencing raft time and dispersal potential of the giant kelp *Macrocystis pyrifera* (Linnaeus) C.Agardh by a combination of tethering experiments and field surveys. The results showed the physiological performance of kelp declined with increasing epiphyte biomass and that higher temperatures increased growth rate of epibionts and fragmentation of kelp. The authors concluded from the results of both the experiments and field surveys that *M. pyrifera* dispersal is dependent on low temperature and moderate irradiance conditions, with high temperatures and higher irradiance reducing overall raft-time and hence dispersal potential [@rothausler2011]. Raft-time was estimated from the results of past studies by @graiff2016 and @graiff2013 for each season. For summer the raft time was 22 days, autumn was 41 days, winter was 38 days, and spring was 33 days.

@@

```{r morphology surface area table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE}
set_flextable_defaults(fonts_ignore=FALSE)
site_surface_area <- set_caption(site_surface_area, "Table 1:Summary of the attributes and estimates used to calculate the overall site-specific surface area of kelp individuals used in the various simulations.") 
#site_surface_area <- set_table_properties(site_surface_area, layout = "autofit")
site_surface_area
```

```{r simulation table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE}
set_flextable_defaults(fonts_ignore=FALSE)
chapter4_table2 <- read_csv2(file = "tables/chapter_4/simulation_table.csv")
chapter4_table2 <- flextable(data = chapter4_table2)
chapter4_table2 <- theme_zebra(chapter4_table2, odd_header = "transparent", even_header = "transparent")
chapter4_table2 <- align(x= chapter4_table2, align = "center", part = "all")
chapter4_table2 <- theme_vanilla(chapter4_table2)
chapter4_table2 <- fontsize(chapter4_table2, size = 8, part = "body")
chapter4_table2 <- fontsize(chapter4_table2, size = 8, part = "header")
chapter4_table2 <- colformat_num(chapter4_table2,digits = 1)
chapter4_table2 <- set_table_properties(chapter4_table2, layout = "autofit")
chapter4_table2 <- set_caption(chapter4_table2, "Table 2: Summary of the simulations run for the study. Simulations were run for each site over the course of 2 years and then pooled based on season for each year. The details pertaining are provided in the below. The start of each year was the beginning of Austral summer.") 
```

@@

To compare virtual kelp trajectories between sites the mean trajectories were calculated, and density distribution maps were produced for each phase of the virtual kelp life cycle (surface, subsurface, beached and sunk). Density distribution maps are an effective method for assessing pathways and ocean connectivity [@vanSebille2018]. To aid in seasonal comparison of trajectories, proportions of each kelp phase at the end of each simulation run was calculated and compared. The different phases were surface, subsurface, beached, sunk and out of bounds.

In addition to the above analyses, the distance particles traveled was also measured and used to produce boxplots to aid in discerning patterns and distributions from the density distribution maps. Distance traveled from the release site can help reveal topographic steering and momentum energy transfer from the wind field to kelp particles [@vanSebille2018]. If the wind direction is in the same direction as the current the virtual particles will cover larger distances due to an increase in the velocity because of momentum energy transfer [@putman2020; @putman2018; @vanSebille2018]. Distance boxplots are presented in the appendix.
